\addcontentsline{toc}{chapter}{Acknowledgements}

\chapter*{Acknowledgements}

First and foremost, I would like to thank my supervisor Dr. Erik Zenner for his continual help and support throughout the project period. Discussions with him have been very fruitful and informative, especially since I do not come from a mathematical background. He gave me a great introduction to various theoretical concepts which were needed for the project, thus providing me a starting point to read and understand them further. I haev been very happy with the way this project shaped up, thanks to Erik. 

I thank two cryptanalysts Karsten Nohl and Marcos el Ruptor, who have always given a great response to my mails asking information about the HiTag2 cipher and feedback on the results of my tradeoff attacks. They have been very helpful through all the information and insight given to me. 

I thank Cristopher Clark for allowing me to use his hashtable implementation. I found his implementation to be very professional, and I admit having learnt certain aspects about program structuring from it and also incorporating them in my implementation. 

I thank my friends in India, Saurabh Bisht and Siddharth Arora, for going through my thesis and pointing out several important mistakes in the draft version.

Importantly, I thank the international office at DTU for providing me a tuition waiver for two years to study computer science at IMM. Without the waiver, I would never have come to DTU in the first place and nor would this project happened. Thanks for believing in me then.

Finally, I thank my loving parents and dear brother for making this project successful. They always supported me with my interest in higher education and believed in me at every point. 